Advances in Artificial Intelligence for Data Visualization: Developing Computer Vision Models to Automate Reading of Data Plots, with Application to Predictive Model Diagnostics

### 2024-07-28 Thesis/03-chap3.qmd
further update for emi's comment

The practice of delivering a residual plot as a lineup is generally regarded as a valuable approach. Beyond its application in residual diagnostics, the lineup protocol has integrated into the analysis of diverse subjects. For instance, Loy and Hofmann [-@loy2013diagnostic; -@loy2014hlmdiag; -@loy2015you] illustrated its applicability in diagnosing hierarchical linear models. Additionally, @widen2016graphical demonstrated its utility in geographical research, while @krishnan2021hierarchical explored its effectiveness in forensic examinations.

A primary limitation of the lineup protocol lies in its reliance on human judgements [see @li2023plot about the practical limitations]. Unlike conventional statistical tests that can be performed computationally in statistical software, the lineup protocol requires human evaluation of images. This characteristic makes it less suitable for large-scale applications, given the associated high labour costs and time requirements. There is a substantial need to develop an approach to substitute these human judgement with an automated reading of data plots using machines.

The utilization of computers to interpret data plots has a rich history, with early efforts such as "Scagnostics" by @tukey1985computer, a set of numerical statistics that summarise features of scatter plots. @wilkinson2005graph expanded on this work, introducing scagnostics based on computable measures applied to planar proximity graphs. These measures, including, but not limited to, "Outlying", "Skinny", "Stringy", "Straight", "Monotonic", "Skewed", "Clumpy", and "Striated" aimed to characterize outliers, shape, density, trend, coherence and other characteristics of the data. While this approach has been inspiring, there is a recognition [@buja2009statistical] that it may not capture all the necessary visual features that differentiate true residual plots from null plots. A more promising alternative entails enabling machines to learn the function for extracting visual features from residual plots. Essentially, this means empowering computers to discern the crucial visual features for residual diagnostics and determining the method to extract them.

@@

```{r fig-false-finding, fig.width = 8, fig.height = 8*4/5, fig.pos="!h", fig.cap="An example residual vs fitted values plot (red line indicates 0). The vertical spread of the data points varies with the fitted values. This often indicates the existence of heteroskedasticity. The Breusch-Pagan test rejects this residual plot at 95\\% significance level ($p\\text{-value} = 0.046$)."}
set.seed(452)
ori_x <- rand_lognormal()
mod <- heter_model(b = 0, x = closed_form(~-ori_x))
ori_dat <- mod$gen(300)

@@

In this study, we chose to define and use a distance between a true residual plot and a theoretical "good" residual plot. @vo2016localizing have also demonstrated that defining a proper distance between images can enhance the matching accuracy in image search compared to a binary outcome model.

@@

To develop a computer vision model for assessing residual plots within the visual inference framework, it is important to precisely define a numerical measure of "difference" or "distance" between plots. This distance can take the form of a basic statistical operation on pixels, such as the sum of square differences. Alternatively, it could involve established image similarity metrics like the Structural Similarity Index Measure [@wang2004image]. The challenge lies in the fact that metrics tailored for image comparison may not be suitable for evaluating data plots, where only essential plot elements require assessment [@chowdhury2018measuring]. Furthermore, scagnostics mentioned in @sec-model-introduction could be used to construct distance measures for residual plots comparison, but the functional form still needs to be carefully refined to accurately reflect the extent of the violations.

In this section, we will introduce a distance measure between a true residual plot and a theoretical 'good' residual plot. This measure quantifies the divergence between the residual distribution of a given fitted regression model and that of a correctly specified model. The computation assumes knowledge of the data generating processes for predictors and response variables. Since these processes are often unknown in practice, we will discuss a method to estimate this distance using a computer vision model in Section @sec-model-distance-estimation.

@@

Suppose the design matrix $\boldsymbol{X}$ has linearly independent columns, the trace of the hat matrix will equal to the number of columns in $\boldsymbol{X}$ denoted as $k$. As a result, the rank of $\boldsymbol{R}$ is $n - k$, and $Q$ is a degenerate multivariate distribution. To capture the characteristics of $Q$, such as moments, we can simulate a large numbers of $\boldsymbol{\varepsilon}$ and transform it to $\boldsymbol{e}$ to get the empirical estimates. For simplicity, in this study, we replaced the variance-covariance matrix of residuals $\text{cov}(\boldsymbol{e}, \boldsymbol{e}) = \boldsymbol{R}\sigma^2\boldsymbol{R}' = \boldsymbol{R}\sigma^2$ with a full-rank diagonal matrix $\text{diag}(\boldsymbol{R}\sigma^2)$, where $\text{diag}(.)$ sets the non-diagonal entries of a matrix to zeros. The resulting distribution for $Q$ is $N(\boldsymbol{0}_n, \text{diag}(\boldsymbol{R}\sigma^2))$.

@@

Define a proper distance between distributions is usually easier than define a proper distance between data plots. Given the true residual distribution $Q$ and the reference residual distribution $P$, we used a distance measure based on Kullback-Leibler divergence [@kullback1951information] to quantify the difference between two distributions

@@

\noindent and $p(.)$ and $q(.)$ are the probability density functions for distribution $P$ and distribution $Q$ respectively.

@@

$$
\begin{aligned}
D_{KL} &= \frac{1}{2}\left(\log\frac{|\text{diag}(\boldsymbol{W})|}{|\text{diag}(\boldsymbol{R}\sigma^2)|} - n + tr(\text{diag}(\boldsymbol{W})^{-1}\text{diag}(\boldsymbol{R}\sigma^2)) + \boldsymbol{\mu}_z'(\text{diag}(\boldsymbol{W}))^{-1}\boldsymbol{\mu}_z\right),
\end{aligned}
$$ {#eq-kl-2}

\noindent where $\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z$, and $\boldsymbol{W} = \boldsymbol{R}\boldsymbol{V}\boldsymbol{R}$. The assumed error variance $\sigma^2$ is set to be $tr(\boldsymbol{V})/n$, which is the expectation of the estimated variance.

@@

In order to approximate $D_{KL}$ in a practically computable manner, the elements of $\boldsymbol{e}$ are assumed to be independent of each other. This assumption solves both of the issues mentioned above. First, we no longer need to integrate over $n$ random variables. The result of @eq-kl-1 is now the sum of the Kullback-Leibler divergence evaluated for each individual residual due to the assumption of independence between observations. Second, it is not required to know the joint probability density $p(\boldsymbol{e})$ any more. Instead, the evaluation of Kullback-Leibler divergence for an individual residual relies on the knowledge of the marginal density $p_i(e_i)$, where $e_i$ is the $i$-th residual for $i = 1, ..., n$. This is much easier to approximate through simulation. It is also worth mentioning that this independence assumption generally will not hold, since $\text{cov}(e_i, e_j) \neq 0$ if $\boldsymbol{R}_{ij} \neq 0$ for any $1 \leq i < j \leq n$, but its existence is essential for reducing the computational cost.

@@

$$
\begin{aligned}
D_{KL} &\approx \sum_{i = 1}^{n} \hat{D}_{KL}^{(i)}, \\
\hat{D}_{KL}^{(i)} &= \frac{1}{m}\sum_{j = 1}^{m} log\frac{\hat{p_i}(B_{ij})}{q(B_{ij})},
\end{aligned} 
$$ {#eq-kl-3}

\noindent where $\hat{D}_{KL}^{(i)}$ is the estimator of the Kullback-Leibler divergence for an individual residual $e_i$, $B_{ij}$ is the $i$-th row and $j$-th column entry of the matrix $B$, $\hat{p_i}(.)$ is the kernel density estimator of $p_i(.)$, $q(.)$ is the normal density function with mean zero and an assumed variance estimated as $\widehat{\sigma^2} = \sum_{b \in vec(B)}(b - \sum_{b \in vec(B)} b/nm)^2/(nm - 1)$, and $vec(.)$ is the vectorization operator which turns a $n \times m$ matrix into a $nm \times 1$ column vector by stacking the columns of the matrix on top of each other.

@@

In the previous sections, we have defined a distance measure given in @eq-kl-0 for quantifying the difference between the true residual distribution $P$ and an ideal reference distribution $Q$. It can be noticed that this distance measure can only be computed when the data generating process is known. In reality, we often have no knowledge about the data generating process, otherwise we do not need to fit a linear regression model in the first place.

@@

$$
\text{MVI} = C + \hat{D} - log(n),
$$ {#eq-mvi}

>>>>

The practice of delivering a residual plot as a lineup is generally regarded as a valuable approach. Beyond its application in residual diagnostics, the lineup protocol has been integrated into the analysis of diverse subjects. For instance, Loy and Hofmann [-@loy2013diagnostic; -@loy2014hlmdiag; -@loy2015you] illustrated its applicability in diagnosing hierarchical linear models. Additionally, @widen2016graphical and @fieberg2024using demonstrated its utility in geographical and ecology research respectively, while @krishnan2021hierarchical explored its effectiveness in forensic examinations.

A practical limitation of the lineup protocol lies in its reliance on human judgements [see @li2023plot about the practical limitations]. Unlike conventional statistical tests that can be performed computationally in statistical software, the lineup protocol requires human evaluation of images. This characteristic makes it less suitable for large-scale applications, given the associated high labour costs and time requirements. There is a substantial need to develop an approach to substitute these human judgement with an automated reading of data plots using machines.

The utilization of computers to interpret data plots has a rich history, with early efforts such as "Scagnostics" by @tukey1985computer, a set of numerical statistics that summarise features of scatter plots. @wilkinson2005graph expanded on this work, introducing scagnostics based on computable measures applied to planar proximity graphs. These measures, including, but not limited to, "Outlying", "Skinny", "Stringy", "Straight", "Monotonic", "Skewed", "Clumpy", and "Striated",  aimed to characterize outliers, shape, density, trend, coherence and other characteristics of the data. While this approach has been inspiring, there is a recognition [@buja2009statistical] that it may not capture all the necessary visual features that differentiate true residual plots from null plots. A more promising alternative entails enabling machines to learn the function for extracting visual features from residual plots. Essentially, this means empowering computers to discern the crucial visual features for residual diagnostics and determining the method to extract them.

@@

```{r fig-false-finding, fig.width = 4, fig.height = 4*4/5, fig.pos="!h", fig.cap="An example residual vs fitted values plot (red line indicates 0). The vertical spread of the data points varies with the fitted values. This often indicates the existence of heteroskedasticity. The Breusch-Pagan test rejects this residual plot at 95\\% significance level ($p\\text{-value} = 0.046$)."}
set.seed(452)
ori_x <- rand_lognormal()
mod <- heter_model(b = 0, x = closed_form(~-ori_x))
ori_dat <- mod$gen(300)

@@

In this study, we chose to define and use a distance between a true residual plot and a theoretical "good" residual plot. This is further explained in @sec-model-distance-between-residual-plots. @vo2016localizing have also demonstrated that defining a proper distance between images can enhance the matching accuracy in image search compared to a binary outcome model.

@@

To develop a computer vision model for assessing residual plots within the visual inference framework, it is important to precisely define a numerical measure of "difference" or "distance" between plots. This distance can take the form of a basic statistical operation on pixels, such as the sum of square differences, however, a pixel-to-pixel comparison makes little sense in comparing residual plots where the main interest would be structural patterns. Alternatively, it could involve established image similarity metrics like the Structural Similarity Index Measure [@wang2004image] which compares images by integrating three perception features of an image: contrast, luminance, and structure (related to average, standard deviation and correlation of pixel values over a window, respectively). These image similarity metrics are tailored for image comparison in vastly different tasks to evaluating data plots, where only essential plot elements require assessment [@chowdhury2018measuring]. We can alternatively define a notion of distance by integrating key plot elements (instead of key perception features like luminance, contrast, and structure), such as those captured by scagnostics mentioned in @sec-model-introduction, but the functional form still needs to be carefully refined to accurately reflect the extent of the violations.

In this section, we introduce a distance measure between a true residual plot and a theoretical 'good' residual plot. This measure quantifies the divergence between the residual distribution of a given fitted regression model and that of a correctly specified model. The computation assumes knowledge of the data generating processes for predictors and response variables. Since these processes are often unknown in practice, we will discuss a method to estimate this distance using a computer vision model in Section @sec-model-distance-estimation.

@@

Suppose the design matrix $\boldsymbol{X}$ has linearly independent columns, the trace of the hat matrix $\boldsymbol{H} = \boldsymbol{X}(\boldsymbol{X}^\top\boldsymbol{X})^{-1}\boldsymbol{X}^\top$ will equal to the number of columns in $\boldsymbol{X}$ denoted as $k$. As a result, the rank of $\boldsymbol{R}$ is $n - k$, and $Q$ is a degenerate multivariate distribution. To capture the characteristics of $Q$, such as moments, we can simulate a large numbers of $\boldsymbol{\varepsilon}$ and transform it to $\boldsymbol{e}$ to get the empirical estimates. For simplicity, in this study, we replaced the variance-covariance matrix of residuals $\text{cov}(\boldsymbol{e}, \boldsymbol{e}) = \boldsymbol{R}\sigma^2\boldsymbol{R}^\top = \boldsymbol{R}\sigma^2$ with a full-rank diagonal matrix $\text{diag}(\boldsymbol{R}\sigma^2)$, where $\text{diag}(.)$ sets the non-diagonal entries of a matrix to zeros. The resulting distribution for $Q$ is $N(\boldsymbol{0}_n, \text{diag}(\boldsymbol{R}\sigma^2))$.

@@

Defining a proper distance between distributions is usually easier than defining a proper distance between data plots. Given the true residual distribution $Q$ and the reference residual distribution $P$, we used a distance measure based on Kullback-Leibler divergence [@kullback1951information] to quantify the difference between two distributions as

@@

\noindent and $p(.)$ and $q(.)$ are the probability density functions for distribution $P$ and distribution $Q$, respectively.

@@

$$
\begin{aligned}
D_{KL} &= \frac{1}{2}\left(\log\frac{|\boldsymbol{W}|}{|\text{diag}(\boldsymbol{R}\sigma^2)|} - n + \text{tr}(\boldsymbol{W}^{-1}\text{diag}(\boldsymbol{R}\sigma^2)) + \boldsymbol{\mu}_z^\top(\boldsymbol{W})^{-1}\boldsymbol{\mu}_z\right),
\end{aligned}
$$ {#eq-kl-2}

\noindent where $\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z$, and $\boldsymbol{W} = \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R})$. The assumed error variance $\sigma^2$ is set to be $tr(\boldsymbol{V})/n$, which is the expectation of the estimated variance.

@@

In order to approximate $D_{KL}$ in a practically computable manner, the elements of $\boldsymbol{e}$ are assumed to be independent of each other. This assumption solves both of the issues mentioned above. First, we no longer need to integrate over $n$ random variables. The result of @eq-kl-1 is now the sum of the Kullback-Leibler divergence evaluated for each individual residual due to the assumption of independence between observations. Second, it is not required to know the joint probability density $p(\boldsymbol{e})$ any more. Instead, the evaluation of Kullback-Leibler divergence for an individual residual relies on the knowledge of the marginal density $p_i(e_i)$, where $e_i$ is the $i$-th residual for $i = 1, ..., n$. This is much easier to approximate through simulation. It is also worth mentioning that this independence assumption generally will not hold if $\text{cov}(e_i, e_j) \neq 0$ for any $1 \leq i < j \leq n$, but its existence is essential for reducing the computational cost.

@@

$$
\begin{aligned}
D_{KL} &\approx \sum_{i = 1}^{n} \hat{D}_{KL}^{(i)}, \\
\hat{D}_{KL}^{(i)} &= \frac{1}{m}\sum_{j = 1}^{m} \log\frac{\hat{p_i}(B_{ij})}{q(B_{ij})},
\end{aligned} 
$$ {#eq-kl-3}

\noindent where $\hat{D}_{KL}^{(i)}$ is the estimator of the Kullback-Leibler divergence for an individual residual $e_i$, $B_{ij}$ is the $i$-th row and $j$-th column entry of the matrix $B$, $\hat{p_i}(.)$ is the kernel density estimator of $p_i(.)$, $q(.)$ is the normal density function with mean zero and an assumed variance estimated as $\widehat{\sigma^2} = \sum_{b \in vec(\boldsymbol{B})}(b - \sum_{b \in vec(\boldsymbol{B})} b/nm)^2/(nm - 1)$, and $vec(.)$ is the vectorization operator which turns a $n \times m$ matrix into a $nm \times 1$ column vector by stacking the columns of the matrix on top of each other.

@@

In the previous sections, we have defined a distance measure given in @eq-kl-0 for quantifying the difference between the true residual distribution $P$ and an ideal reference distribution $Q$. It can be noticed that this distance measure can only be computed when the data generating process is known. In reality, we often have no knowledge about the data generating process, otherwise we do not need to do a residual diagnostic in the first place.

@@

$$
\text{MVI} = C + \hat{D} - \log(n),
$$ {#eq-mvi}
